---
title: "Getting started with edetect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with edetect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r setup}
library(edetect)
```

## What an e-detector is

A control chart watches a stream of observations and raises an alarm when the
process seems to have changed. The classical charts (Shewhart, CUSUM, EWMA) are
calibrated under a model for the in-control process, usually Gaussian, and their
false-alarm rate is only as good as that model.

An **e-detector** (Shin, Ramdas and Rinaldo, 2024) replaces the model by a
*class*: everything the user is willing to assume about the process before the
change, for example "observations lie in [0, 1] and their mean is at most 0.5".
For every distribution in the class, the expected time to a false alarm is at
least the target **average run length** (ARL). The detector accumulates
*e-values*, bets against the pre-change class, started at every time point;
the Shiryaev--Roberts type sums them, the CUSUM type keeps the largest, and the
alarm is raised when the accumulated evidence reaches the ARL.

The price of the guarantee is paid in detection delay: the detector cannot
know the size of the change, so it bets on a mixture of shifts.

## A bounded stream

A loss in [0, 1] whose mean should stay at 0.5. After 60 observations the mean
rises to 0.7.

```{r bounded}
set.seed(1)
x <- c(runif(60, 0.3, 0.7), runif(60, 0.5, 0.9))
chart <- edetect_x(x, bounds = c(0, 1), center = 0.5, arl = 370, direction = "up")
chart
edetect_alarm(chart)
autoplot(chart)
```

The detector stopped at the alarm; the observations after it were not consumed
and a warning said so. `edetect_report()` lists what was assumed, how the
detector was built and what happened, and `tidy()` returns the trajectory.

```{r report}
edetect_report(chart)
head(tidy(chart))
```

## Phase I and Phase II

The centre is usually not known but estimated from in-control data. Supplying
`phase1` sets it and records the source as an assumption of the chart.

```{r phase1}
design <- edetect_design(arl = 370, class = "bounded", bounds = c(0, 1),
                         phase1 = runif(100, 0.3, 0.7), direction = "both")
design
```

## Attribute charts

A p chart monitors the proportion of nonconforming units in subgroups of size
`n`; a c chart counts nonconformities per unit; a u chart counts them per unit
of exposure. The pre-change class for p is "each trial has success probability
at most the centre" (dependence within a subgroup is allowed); for c and u it
is "counts are sub-Poisson with the declared rate", which excludes
overdispersion.

```{r attributes}
set.seed(2)
defects <- c(rbinom(40, 100, 0.02), rbinom(30, 100, 0.06))
p_chart <- edetect_p(defects, n = 100, center = 0.02, arl = 370)
edetect_alarm(p_chart)
autoplot(p_chart)
```

With the daily COVID-19 deaths of Recife from the `shewhartr` package, a c chart
with the first four weeks as Phase I asks when the count left its early level.

```{r recife, eval = requireNamespace("shewhartr", quietly = TRUE)}
deaths <- shewhartr::cvd_recife$new_deaths
c_chart <- edetect_c(deaths[-(1:28)], phase1 = deaths[1:28], arl = 370, direction = "up")
edetect_alarm(c_chart)
```

## Regression chart

When the in-control level depends on covariates, a Phase I model is fitted and
the Phase II residuals are monitored as a sub-Gaussian stream centred at zero
with the residual scale of the fit.

```{r regression}
set.seed(3)
d1 <- data.frame(load = 1:80); d1$time <- 10 + 0.4 * d1$load + rnorm(80, sd = 1)
d2 <- data.frame(load = 81:140); d2$time <- 10 + 0.4 * d2$load + rnorm(60, sd = 1) + c(rep(0, 30), rep(2.5, 30))
r_chart <- edetect_regression(time ~ load, phase1 = d1, phase2 = d2, arl = 370, direction = "up")
edetect_alarm(r_chart)
```

## Checking the guarantee and the delay

`edetect_arl()` simulates run lengths of a design: without a change it
estimates the ARL (which must be at least the target), with a change it
estimates the detection delay.

```{r arl}
design <- edetect_design(arl = 100, class = "subgaussian", center = 0, sigma = 1, direction = "up")
set.seed(4)
edetect_arl(design, pre = function(n) rnorm(n), n_rep = 100, max_t = 3000)$summary
edetect_arl(design, pre = function(n) rnorm(n), post = function(n) rnorm(n, 1),
            change_at = 50, n_rep = 100, max_t = 500)$summary
```

## References

Shin, J., Ramdas, A. and Rinaldo, A. (2024). E-detectors: a nonparametric
framework for sequential change detection. *The New England Journal of
Statistics in Data Science*, 2(2), 229--260.
